DisReserv (v1.0): Analytical displacement solution due to reservoir compaction with arbitrary geometry and under arbitrary pressure changes

### 2020-09-17 manuscript/manuscript-reservoir.tex

\cite{Geertsma&Opstal73} applied the nucleus-of-strain concept in the half-space 
to calculate the spatial subsidence distribution due to the production of reservoir with an arbitrary 3D shape.
By assuming a producing reservoir embedded in an a homogeneous, isotropic, and elastic medium, and a reservoir model in which the pressure perturbations are related to the displacement field by a linear relationship, \cite{Geertsma&Opstal73} discretized the reservoir into a grid of  prisms and calculated the displacement due to the pressure change in the whole reservoir  by the superposition of the displacement due to the constant pressure change in each prism.
\cite{Tempone10} adopted the same reservoir model used in \cite{Geertsma&Opstal73} and 
extended the nucleus-of-strain concept in the half-space to consider the effects of a rigid basement. 
Similarly, \cite{Tempone10} assumed a reservoir embedded in an a homogeneous, isotropic, and elastic medium and calculated the displacement, stress and strain fields subject to uniform depletion.
The main drawbacks in \cite{Geertsma&Opstal73} and \cite{Tempone10} are the assumption of homogeneous reservoir and the solution is only valid outside the reservoir.
In these case, the displacements within the reservoir are calculated by a linear interpolation of the displacements at the upper and lower edges of the reservoir \citep{Tempone12}.

@@

The present work assumes a linear elastic semi-infinite medium and provides an analytical solution for displacement field due to an arbitrarily-shaped reservoir under arbitrary distribution of pressure changes. 
Like \cite{Munoz&Roehl17} we used the nucleus-of-strain concept and discretized the
reservoir into a grid of 3D prisms along the $x-$, $y-$ and $-z$directions.
We also consider the nucleus of strain as an infinitesimal volume element  for each prism and the displacement solution due to each prism is obtained by a three-dimensional integration over the prism volume. 
The final displacement field due to the whole reservoir is the sum of the displacements produced by the prisms. 
In contrast with \cite{Munoz&Roehl17}'s method we take advantage the similarity between 
the  equations for calculating the displacement field due to a volume source in a half-space under a pressure variation and the gravitational potential due to a volume source under a density variation.
This similarity makes possible the use of closed expressions of the gravitational potential and its derivatives produced by the 3D right rectangular prism derived by \cite{Nagyetal2000} and (\citeyear{Nagyetal2002}) and \cite{Fukushima2020}  for calculating the displacement field due to a volume source under a pressure variation.
The adopted exact analytical formulae of the gravitational field  are valid expressions either outside or inside the prisms because the implemented expressions make use of  modified arctangent function proposed by \cite{Fukushima2020}. 
We present routines written in Python language (Python 3.7.6)  to calculate the displacement fields due to a reservoir with arbitrary shape and distribution of pressure changes. 
We validate our equations by verifying that at the free surface the stress fields are null.
Tests with synthetic data validate our approach.

@@

From the potential theory, a particular  solution of equation \ref{eq:poisson} is
\begin{equation}
\phi(x,y,z) = -  \frac{m}{4 \pi} \int\int\limits_{v}\int \frac{T(x^{\prime}, y^{\prime}, z^{\prime} )}{\sqrt{(x - x^{\prime})^{2} + (y - y^{\prime})^{2} + (z - z^{\prime})^{2}}} \: dv,
\label{eq:phi}
\end{equation}
where $\phi(x,y,z)$ represents the Newtonian gravitational potential \citep{Kellogg29} at the coordinates $x, y$ and $z$ produced by a continuous distribution of mass defined by a set of very small masses $ -  \frac{m}{4 \pi}  \: T( x^{\prime}, y^{\prime}, z^{\prime} ) \: dv$, where $T(x^{\prime}, y^{\prime}, z^{\prime}) $ stands for the density distribution. 
The integral in equation \ref{eq:phi} is conducted over the coordinates $x^{\prime}$,$y^{\prime}$ and $z^{\prime}$, denoting, respectively, the x-, y-, and z-coordinates of an arbitrary point belonging to the interior of the volume $v$ of the gravity source.

@@

We validate our equations by verifying if the null stress acting through the free surface (equation \ref{eq:null_stress}) is satisfied.
Hence, by taking the elements of stress vectors $\mbox{\boldmath$\sigma_{1}$}_{i}$ and 
$\mbox{\boldmath$\sigma_{2}$}_{i}$  at the $i$th coordinates of the free surface ($x_i$, $y_i$ and $z_i = 0$) the following relationship:
\begin{equation}
{\widehat{xz}_{1}} + {\widehat{xz}_{2}} = {\widehat{yz}_{1}} + {\widehat{yz}_{2}}
+ {\widehat{zz}_{1}} + {\widehat{zz}_{2}}  = 0
\label{eq:null_stress_i}
\end{equation}
must be met.

@@

From equation  \ref{eq:nucleus-u2-vector}, we can get the $ x-$ and $y-$components of the displacement vectors ${\bf{u_{2}}}_{i}$  (the integrand of the second integral of equation \ref{eq:nucleus-ui_M_prisms})
\begin{equation}
{u_{2}}_\alpha =  \frac{A  \: (1 + \nu)}{E} \: \Bigg[ 
(3  - 4 \nu)    \: \frac{\partial }{\partial \alpha} {\frac{1}{{R_2}_{ij}}}\: 
+ 2 \: z_{i} \:  
\frac{\partial^{2}  }{\partial \alpha \partial z} {\frac{1}{{R_2}_{ij}}}  \: \Bigg]
\Delta p_{j} ,
\label{eq:u2_alpha_component}
\end{equation}
where  $\alpha$  belongs to the set of $ x-$ and $y-$directions, and the $z-$component of the displacement vectors ${\bf{u_{2}}}_{i}$  
\begin{equation}
{u_{2}}_z =  \frac{A  \: (1 + \nu)}{E} \: \Bigg[ 
(3  - 4 \nu)    \: - \frac{\partial }{\partial z} {\frac{1}{{R_2}_{ij}}}\: 
+ 2 \: z_{i} \:  
\frac{\partial^{2}  }{\partial z^{2}} {\frac{1}{{R_2}_{ij}}}  \: \Bigg]
\Delta p_{j}
\label{eq:u2_z_component}
\end{equation}

@@

To apply our methodology, we  discretized the cylinder  along the $x-$ and $y-$ directions into an $20 \times 20$ grid of prisms. Hence, we totalized 400 prisms all of them centered at 800 m deep, with depths to the top and to the bottom at 750 m and 850 m  and with pressure contrast $\Delta p_j$, $j = 1, ..., 400$ equal to $-10$ MPa.
To apply the Geertsma’s method \citep{Geertsma73}, we used the disk-shaped reservoir \citep{Fjaer08} with dimensions and physical properties described above.

@@

We  discretized the cylinder  along the $x-$, $y-$ and $z-$ directions into an $20 \times 20 \times 2$ grid of prisms.
This simulation totalized 800 prisms whose thicknesses are 50 m.
The 400 deepest prisms are centered at 825 deep, with pressure contrast equal to $-20$ MPa and the 400 shalowest prisms are centered at 775 deep, with pressure contrast equal to $-40$ MPa

@@

In this numerical application, the reservoir model is a simplification of a realistic reservoir located in a production oil field in offshore Brazil.
The entire reservoir model comprises dimensions of 14 km in the north-axis, 13 km in the
east-axis, and 0.6 km in the down-axis. 
The depths to the top and bottom of the reservoir model are 2,712 m and 3,312 m, respectively. 
The components of the displacements are calculated at 0 m deep, 
on a regular grid of 100 $\times$ 80  observation points along the north- and east-directions, respectively. 
We  discretized the reservoir  along the $x-$, $y-$ and $z-$ directions into an $14 \times 13 \times 2$ grid of prisms.
The Young’s modulus is  3300 (in MPa), the Poisson's coefficient is 0.25, and
the uniaxial compaction coefficient $C_{m}$  (equation \ref{eq:Cm}) is $2.2525 \: 10^{-4}$
$\textrm{ MPa}^{-1}$.
In the 3D perspective view of the pore pressure distribution of the simulated reservoir shown in Figure \ref{fig:pressure_complex_reservoir}, we can see that the pressures vary from 0 to -0.72 MPa
Figure \ref{fig:displacement_complex_reservoir} shows cross-sections at $x  = 8$ km of the horizontal and vertical displacements, in 2D contour plots, calculated in the whole reservoir by using our methodology.
Figure \ref{fig:Null_stress_complex_reservoir} shows the  null stress through the free surface (equations \ref{eq:null_stress} and \ref{eq:null_stress_i}) due to reservoir with arbitrary geometry and under arbitrary pressure disribution.

@@

\authorcontribution{} %% this section is mandatory

@@

\begin{acknowledgements}
Valeria C.F. Barbosa was supported by fellowships from: CNPq (grant 307135/2014- 4) and FAPERJ (grant E-26/202.582/2019). 
Vanderlei C. Oliveira Jr. was supported by fellowships from: CNPq (grant 308945/2017-4) and FAPERJ (grant E-26/202.729/2018). 
\end{acknowledgements}

>>>>

\cite{Geertsma&Opstal73} applied the nucleus-of-strain concept in the half-space 
to calculate the spatial subsidence distribution due to the production of reservoir with an arbitrary 3D shape.
By assuming a producing reservoir embedded in an a homogeneous, isotropic, and elastic medium, and a reservoir model in which the pressure perturbations are related to the displacement field by a linear relationship, \cite{Geertsma&Opstal73} discretized the reservoir into a grid of  prisms and calculated the displacement due to the pressure change in the whole reservoir  by the superposition of the displacement due to the constant pressure change in each prism.
\cite{Tempone10} adopted the same reservoir model used in \cite{Geertsma&Opstal73} and 
extended the nucleus-of-strain concept in the half-space to consider the effects of a rigid basement. 
Similarly, \cite{Tempone10} assumed a reservoir embedded in an a homogeneous, isotropic, and elastic medium and calculated the displacement, stress and strain fields subject to uniform depletion.
The main drawbacks in \cite{Geertsma&Opstal73} and \cite{Tempone10} are the assumption of homogeneous reservoir and the solution is only valid outside the reservoir.
In this case, the displacements within the reservoir are calculated by a linear interpolation of the displacements at the upper and lower edges of the reservoir \citep{Tempone12}.

@@

The present work assumes a linear elastic semi-infinite medium and provides an analytical solution for displacement field due to an arbitrarily-shaped reservoir under arbitrary distribution of pressure changes. 
Like \cite{Munoz&Roehl17} we used the nucleus-of-strain concept and discretized the
reservoir into a grid of 3D prisms along the $x-$, $y-$ and $-z$ directions.
We also consider the nucleus of strain as an infinitesimal volume element  for each prism and the displacement solution due to each prism is obtained by a three-dimensional integration over the prism volume. 
The final displacement field due to the whole reservoir is the sum of the displacements produced by the prisms. 
In contrast with \cite{Munoz&Roehl17}'s method we take advantage the similarity between 
the  equations for calculating the displacement field due to a volume source in a half-space under a pressure variation and the gravitational potential due to a volume source under a density variation.
This similarity makes possible the use of closed expressions of the gravitational potential and its derivatives produced by the 3D right rectangular prism derived by \cite{Nagyetal2000} and (\citeyear{Nagyetal2002}) and \cite{Fukushima2020}  for calculating the displacement field due to a volume source under a pressure variation.
The adopted exact analytical formulae of the gravitational field  are valid expressions either outside or inside the prisms because the implemented expressions make use of  modified arctangent function proposed by \cite{Fukushima2020}. 
We present routines written in Python language (Python 3.7.6)  to calculate the displacement fields due to a reservoir with arbitrary shape and distribution of pressure changes. 
We validate our equations by verifying that at the free surface the stress fields are null.
Tests with synthetic data validate our approach.

@@

From the potential theory, a particular  solution of equation \ref{eq:poisson} is
\begin{equation}
\phi(x,y,z) = -  \frac{m}{4 \pi} \int\int\limits_{v}\int \frac{T(x^{\prime}, y^{\prime}, z^{\prime} )}{\sqrt{(x - x^{\prime})^{2} + (y - y^{\prime})^{2} + (z - z^{\prime})^{2}}} \: dv,
\label{eq:phi}
\end{equation}
where $\phi(x,y,z)$ represents the Newtonian gravitational potential \citep{Kellogg29} at the coordinates $x, y$ and $z$ produced by a continuous distribution of mass defined by a set of very small masses $ -  \frac{m}{4 \pi}  \: T( x^{\prime}, y^{\prime}, z^{\prime} ) \: dv$, where $T(x^{\prime}, y^{\prime}, z^{\prime}) $ stands for the density distribution. 
The integral in equation \ref{eq:phi} is conducted over the coordinates $x^{\prime}$,$y^{\prime}$ and $z^{\prime}$, denoting, respectively, the $x-$, $y-$ and $-z$ coordinates of an arbitrary point belonging to the interior of the volume $v$ of the gravity source.

@@

We validate our equations by verifying if the null stress acting through the free surface (equation \ref{eq:null_stress}) is satisfied.
Hence, by taking the elements of stress vectors $\mbox{\boldmath$\sigma_{1}$}_{i}$ and 
$\mbox{\boldmath$\sigma_{2}$}_{i}$  at the $i$th coordinates of the free surface ($x_i$, $y_i$ and $z_i = 0$) the following relationship:
\begin{equation}
{\widehat{xz}_{1}} + {\widehat{xz}_{2}} 
= {\widehat{yz}_{1}} + {\widehat{yz}_{2}}
= {\widehat{zz}_{1}} + {\widehat{zz}_{2}}  
= 0
\label{eq:null_stress_i}
\end{equation}
must be met.

@@

From equation  \ref{eq:nucleus-u2-vector}, we can get the $ x-$ and $y-$components of the displacement vectors ${\bf{u_{2}}}_{i}$  (the integrand of the second integral of equation \ref{eq:nucleus-ui_M_prisms})
\begin{equation}
{u_{2}}_\alpha =  \frac{A  \: (1 + \nu)}{E} \: \Bigg[ 
(3  - 4 \nu)    \: \frac{\partial }{\partial \alpha} {\frac{1}{{R_2}_{ij}}}\: 
+ 2 \: z_{i} \:  
\frac{\partial^{2}  }{\partial \alpha \partial z} {\frac{1}{{R_2}_{ij}}}  \: \Bigg]
\Delta p_{j} ,
\label{eq:u2_alpha_component}
\end{equation}
where  $\alpha$  belongs to the set of $ x-$ and $y-$directions, and the $z-$component of the displacement vectors ${\bf{u_{2}}}_{i}$  
\begin{equation}
{u_{2}}_z =  \frac{A  \: (1 + \nu)}{E} \: \Bigg[ 
\: - (3  - 4 \nu)  \frac{\partial }{\partial z} {\frac{1}{{R_2}_{ij}}}\: 
+ 2 \: z_{i} \:  
\frac{\partial^{2}  }{\partial z^{2}} {\frac{1}{{R_2}_{ij}}}  \: \Bigg]
\Delta p_{j}
\label{eq:u2_z_component}
\end{equation}

@@

To apply our methodology, we  discretized the cylinder  along the $x-$ and $y-$ directions into an $20 \times 20$ grid of prisms. Hence, we totalized 400 prisms all of them centered at 800 m deep, with depths to the top and to the bottom at 750 m and 850 m  and with pressure contrast $\Delta p_j$, $j = 1, ..., 400$ equal to $-10$ MPa.
To apply the Geertsma’s method \citep{Geertsma73}, we used the disk-shaped reservoir described in \cite{Fjaer08} with dimensions and physical properties defined above.

@@

We  discretized the cylinders  along the $x-$, $y-$ and $z-$ directions into an $20 \times 20 \times 2$ grid of prisms.
This simulation totalized 800 prisms whose thicknesses are 50 m.
The 400 deepest prisms are centered at 825 deep, with pressure contrast equal to $-20$ MPa and the 400 shalowest prisms are centered at 775 deep, with pressure contrast equal to $-40$ MPa

@@

In this numerical application, the reservoir model is a simplification of a realistic reservoir located in a production oil field in offshore Brazil.
The entire reservoir model comprises dimensions of 14 km in the north-axis, 13 km in the
east-axis, and 0.6 km in the down-axis. 
The depths to the top and bottom of the reservoir model are 2,712 m and 3,312 m, respectively. 
The components of the displacements are calculated at 0 m deep, 
on a regular grid of 100 $\times$ 80  observation points along the north- and east-directions, respectively. 
We  discretized the reservoir  along the $x-$, $y-$ and $z-$ directions into an $14 \times 13 \times 2$ grid of prisms.
The Young’s modulus is  3300 (in MPa), the Poisson's coefficient is 0.25, and
the uniaxial compaction coefficient $C_{m}$  (equation \ref{eq:Cm}) is $2.2525 \: 10^{-4}$
$\textrm{ MPa}^{-1}$.
In the 3D perspective view of the pore pressure distribution of the simulated reservoir shown in Figure \ref{fig:pressure_complex_reservoir}, we can see that the pressures vary from $0$ to $-0.72$ MPa.
Figure \ref{fig:displacement_complex_reservoir} shows cross-sections at $x  = 8$ km of the horizontal and vertical displacements, in 2D contour plots, calculated in the whole reservoir by using our methodology.
Figure \ref{fig:Null_stress_complex_reservoir} shows the  null stress through the free surface (equations \ref{eq:null_stress} and \ref{eq:null_stress_i}) due to reservoir with arbitrary geometry and under arbitrary pressure disribution.

@@

\authorcontribution{
VCFB and VCOJr deduced the equations in this paper.
VCFB  and ADA implemented the numerical applications presented in this paper.
Most of the python codes were developed by VCOJr with some participation of VCFB.
All authors contributed to the writing of the paper.
} %% this section is mandatory

@@

\begin{acknowledgements}
Andre D. Arelaro thanks Petrobras, especially, the Multiphysics Group.
Valeria C.F. Barbosa was supported by fellowships from: CNPq (grant 307135/2014- 4) and FAPERJ (grant E-26/202.582/2019). 
Vanderlei C. Oliveira Jr. was supported by fellowships from: CNPq (grant 308945/2017-4) and FAPERJ (grant E-26/202.729/2018). 
\end{acknowledgements}
